---
title: "Getting started with gipsDA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gipsDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`gipsDA` extends classical Linear Discriminant Analysis (LDA) and
Quadratic Discriminant Analysis (QDA) by replacing standard empirical covariance
estimates with covariance estimates regularized by permutation-invariant
structures learned with the `gips` framework.

The package is especially useful when the covariance matrix is difficult to
estimate reliably, for example when the number of features is large relative to
the number of observations. In such settings, classical covariance estimators may
be unstable or singular.

The package provides three main model families:

- `gipslda()` - LDA with a projected pooled covariance matrix.
- `gipsqda()` - QDA with separately projected class-specific covariance matrices.
- `gipsmultqda()` - QDA with class-specific covariance matrices and a jointly
  estimated permutation structure.

All three functions follow the familiar conventions of `MASS::lda()` and
`MASS::qda()`. They support formula, matrix, and data-frame interfaces, and their
`predict()` methods return class predictions and posterior probabilities.

## Installation

Install the released version from CRAN:

```{r, eval = FALSE}
install.packages("gipsDA")
```

Or install the development version from GitHub:

```{r, eval = FALSE}
pak::pkg_install("AntoniKingston/gipsDA")
```

Load the package:

```{r}
library(gipsDA)
```

## Example data

We use the built-in `iris` data set. To make the example closer to a standard
classification workflow, we create a stratified train/test split.

```{r}
set.seed(42)

train_id <- unlist(
  lapply(split(seq_len(nrow(iris)), iris$Species), sample, size = 35),
  use.names = FALSE
)

train <- iris[train_id, ]
test <- iris[-train_id, ]

table(train$Species)
table(test$Species)
```

## Main functions

The main model-fitting functions are:

| Function | Classical analogue | Covariance structure |
|---|---|---|
| `gipslda()` | `MASS::lda()` | one projected pooled covariance matrix |
| `gipsqda()` | `MASS::qda()` | one projected covariance matrix per class |
| `gipsmultqda()` | `MASS::qda()` | class-specific covariance matrices with one jointly estimated permutation structure |

The package also provides several S3 methods.

| Method | Objects | Purpose |
|---|---|---|
| `predict()` | `gipslda`, `gipsqda`, `gipsmultqda` | class prediction and posterior probabilities |
| `print()` | `gipslda`, `gipsqda`, `gipsmultqda` | compact model summary |
| `coef()` | `gipslda` | LDA coefficients |
| `plot()` | `gipslda` | LDA diagnostic plot |
| `pairs()` | `gipslda` | pairwise LDA visualization |
| `model.frame()` | `gipslda`, `gipsqda`, `gipsmultqda` | extract the model frame when available |

## Model-fitting arguments

The fitting functions follow the standard R modeling style. Depending on the
interface, the model can be fitted using a formula, a matrix, or a data frame.

### Formula interface arguments

| Argument | Meaning |
|---|---|
| `formula` | model formula, for example `Species ~ .` |
| `data` | data frame containing predictors and the grouping variable |
| `subset` | optional subset of rows used for fitting |
| `na.action` | handling of missing values |
| `...` | additional arguments passed to the corresponding lower-level method |

### Matrix and data-frame interface arguments

| Argument | Meaning |
|---|---|
| `x` | matrix or data frame of predictors |
| `grouping` | class labels |
| `...` | additional arguments passed to the default method |

### Main statistical and optimization arguments

| Argument | Used in | Meaning |
|---|---|---|
| `prior` | `gipslda()`, `gipsqda()`, `gipsmultqda()` | prior probabilities of classes |
| `tol` | mainly `gipslda()` | numerical tolerance used during fitting |
| `MAP` | all models | whether to use the single MAP permutation or posterior-weighted averaging |
| `optimizer` | all models | permutation-search strategy, usually `"BF"` or `"MH"` |
| `max_iter` | all models | number of iterations for the Metropolis-Hastings optimizer |
| `weighted_avg` | `gipslda()` only | covariance pooling strategy before projection |

The most important `gipsDA`-specific arguments are `MAP`, `optimizer`,
`max_iter`, and, for `gipslda()`, `weighted_avg`.

## `MAP`: single best permutation vs posterior averaging

The `MAP` argument controls how the permutation-invariant covariance estimate is
constructed.

When `MAP = TRUE`, the model finds the most probable permutation structure and
projects the covariance matrix onto the space invariant under that structure.

```{r}
lda_map <- gipslda(
  Species ~ .,
  data = train,
  MAP = TRUE,
  optimizer = "BF"
)

lda_map
```

When `MAP = FALSE`, the model uses posterior probabilities over retained
permutations and computes a posterior-weighted covariance projection.

```{r}
lda_avg <- gipslda(
  Species ~ .,
  data = train,
  MAP = FALSE,
  optimizer = "BF"
)

lda_avg
```

Use `MAP = TRUE` when you want a single selected structure. Use `MAP = FALSE`
when you want to account for uncertainty in the selected permutation structure.

## `optimizer`: brute force and Metropolis-Hastings

The `optimizer` argument controls how the permutation structure is searched.

| Value | Meaning | Recommended use |
|---|---|---|
| `"BF"` | brute-force search | small number of features |
| `"MH"` | Metropolis-Hastings search | larger number of features |

Brute force is deterministic and searches the relevant permutation space
exhaustively, but its cost grows quickly with the number of features.

```{r}
lda_bf <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)

lda_bf
```

For larger feature spaces, use `"MH"` and control runtime with `max_iter`.

```{r, eval = FALSE}
lda_mh <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "MH",
  max_iter = 1000
)
```

## `max_iter`

`max_iter` controls the number of Metropolis-Hastings iterations when
`optimizer = "MH"`.

```{r, eval = FALSE}
fit <- gipsqda(
  Species ~ .,
  data = train,
  optimizer = "MH",
  max_iter = 1000
)
```

Increasing `max_iter` may improve exploration of the permutation space, but also
increases runtime.

For `optimizer = "BF"`, `max_iter` is not the main computational control.

## `prior`

The `prior` argument specifies class prior probabilities.

By default, class priors are estimated from the training data. You can override
them manually.

```{r}
equal_prior <- rep(1 / length(levels(train$Species)), length(levels(train$Species)))
names(equal_prior) <- levels(train$Species)

lda_prior <- gipslda(
  Species ~ .,
  data = train,
  prior = equal_prior,
  optimizer = "BF"
)

lda_prior$prior
```

The prior vector should have one value per class and should sum to one.

## `weighted_avg` in `gipslda()`

`weighted_avg` is specific to `gipslda()` and controls how the pooled covariance
matrix is constructed before applying the `gips` projection.

```{r}
lda_classic <- gipslda(
  Species ~ .,
  data = train,
  weighted_avg = FALSE,
  optimizer = "BF"
)

lda_weighted <- gipslda(
  Species ~ .,
  data = train,
  weighted_avg = TRUE,
  optimizer = "BF"
)
```

Use:

- `weighted_avg = FALSE` for the classic LDA-style pooled covariance estimator.
- `weighted_avg = TRUE` for the class-size-weighted averaging variant.

## Linear discriminant analysis with `gipslda()`

The formula interface is the most convenient way to fit a model.

```{r}
lda_fit <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)

lda_fit
```

Prediction returns class labels and posterior probabilities.

```{r}
lda_pred <- predict(lda_fit, test)

head(lda_pred$class)
head(lda_pred$posterior)

mean(lda_pred$class == test$Species)
```

## Prediction methods for `gipslda()`

For `gipslda()` objects, prediction can be performed with different methods.

```{r}
pred_plugin <- predict(lda_fit, test, method = "plug-in")
pred_predictive <- predict(lda_fit, test, method = "predictive")
pred_debiased <- predict(lda_fit, test, method = "debiased")

head(pred_plugin$class)
head(pred_predictive$class)
head(pred_debiased$class)
```

The available prediction methods are:

| Method | Meaning |
|---|---|
| `"plug-in"` | standard plug-in discriminant rule |
| `"predictive"` | predictive rule |
| `"debiased"` | debiased rule |

## Quadratic discriminant analysis with `gipsqda()`

`gipsqda()` estimates a separate projected covariance matrix for each class.

```{r}
qda_fit <- gipsqda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)

qda_fit
```

Prediction works analogously to classical QDA.

```{r}
qda_pred <- predict(qda_fit, test)

head(qda_pred$class)
head(qda_pred$posterior)

mean(qda_pred$class == test$Species)
```

## Joint quadratic discriminant analysis with `gipsmultqda()`

`gipsmultqda()` is an intermediate model between `gipslda()` and `gipsqda()`.

Like QDA, it allows each class to have its own covariance matrix. Unlike
`gipsqda()`, it estimates one shared permutation structure for all class
covariance matrices. This is useful when classes may differ in scale or variance,
but are expected to share a common dependency pattern between features.

```{r}
joint_qda_fit <- gipsmultqda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)

joint_qda_fit
```

Prediction returns class labels and posterior probabilities.

```{r}
joint_qda_pred <- predict(joint_qda_fit, test)

head(joint_qda_pred$class)
head(joint_qda_pred$posterior)

mean(joint_qda_pred$class == test$Species)
```

## Comparing the three models

```{r}
acc <- c(
  gipslda = mean(lda_pred$class == test$Species),
  gipsqda = mean(qda_pred$class == test$Species),
  gipsmultqda = mean(joint_qda_pred$class == test$Species)
)

acc
```

## Leave-one-out prediction for QDA models

For `gipsqda()` and `gipsmultqda()` objects, leave-one-out cross-validation can
be requested by omitting `newdata` and using `method = "looCV"`.

In this mode, each training observation is classified as if it had been left out
when fitting the model. This gives an internal estimate of classification
accuracy without creating a separate test set.

```{r}
qda_loo <- predict(qda_fit, method = "looCV")
joint_qda_loo <- predict(joint_qda_fit, method = "looCV")

c(
  gipsqda_loo_accuracy = mean(qda_loo$class == train$Species),
  gipsmultqda_loo_accuracy = mean(joint_qda_loo$class == train$Species)
)
```

## Formula interface

The formula interface is usually the easiest one to use.

```{r}
fit_formula <- gipslda(
  Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = train,
  optimizer = "BF"
)

predict(fit_formula, test)$class[1:10]
```

Formula methods also support standard arguments such as `subset` and `na.action`
through the usual R model-frame workflow.

```{r}
fit_subset <- gipslda(
  Species ~ .,
  data = iris,
  subset = Species != "setosa",
  optimizer = "BF"
)

fit_subset
```

## Matrix interface

The matrix interface separates predictors from class labels.

```{r}
x <- as.matrix(iris[, 1:4])
grouping <- iris$Species

fit_matrix <- gipslda(
  x,
  grouping,
  optimizer = "BF"
)

predict(fit_matrix, x[1:5, ])$class
```

The same interface can be used with `gipsqda()`.

```{r}
qda_matrix <- gipsqda(
  x,
  grouping,
  optimizer = "BF"
)

predict(qda_matrix, x[1:5, ])$class
```

And with `gipsmultqda()`.

```{r}
joint_qda_matrix <- gipsmultqda(
  x,
  grouping,
  optimizer = "BF"
)

predict(joint_qda_matrix, x[1:5, ])$class
```

## Data-frame interface

You can also pass predictors as a data frame and labels separately.

```{r}
x_df <- iris[, 1:4]
y <- iris$Species

fit_df <- gipslda(
  x_df,
  y,
  optimizer = "BF"
)

predict(fit_df, x_df[1:5, ])$class
```

The data-frame interface is also available for QDA-type models.

```{r}
qda_df <- gipsqda(
  x_df,
  y,
  optimizer = "BF"
)

joint_qda_df <- gipsmultqda(
  x_df,
  y,
  optimizer = "BF"
)

predict(qda_df, x_df[1:5, ])$class
predict(joint_qda_df, x_df[1:5, ])$class
```

## Fitted objects

Fitted objects contain the usual discriminant-analysis components, such as class
priors, class counts, group means, and scaling information.

```{r}
names(lda_fit)
names(qda_fit)
names(joint_qda_fit)
```

`gipsDA` objects additionally store optimization information.

```{r}
lda_fit$optimization_info
qda_fit$optimization_info
joint_qda_fit$optimization_info
```

When `MAP = TRUE`, this information corresponds to the selected permutation
structure. When `MAP = FALSE`, it contains posterior-probability information used
for averaging.

## Inspecting fitted model objects

Fitted `gipsDA` models are standard R objects. They contain the usual
discriminant-analysis components, such as class priors, class counts, group
means, scaling information, and additional information about the permutation
optimization.

The most readable way to inspect a fitted model is to print it.

```{r}
print(lda_fit)
```

```{r}
print(qda_fit)
```

```{r}
print(joint_qda_fit)
```

The printed output shows the main fitted quantities, including prior
probabilities, group means, and information about the permutation structures
selected or averaged by the model.

For a more structural view, use `summary()`.

```{r}
summary(lda_fit)
summary(qda_fit)
summary(joint_qda_fit)
```

Because fitted models are list-like S3 objects, it is also useful to inspect
their component names directly.

```{r}
names(lda_fit)
names(qda_fit)
names(joint_qda_fit)
```

The most important components are usually:

| Component | Meaning |
|---|---|
| `prior` | prior probabilities of classes |
| `counts` | number of observations in each class |
| `means` | class-wise feature means |
| `scaling` | scaling or decomposition information used for prediction |
| `ldet` | log-determinant information for QDA-type models |
| `lev` | class labels |
| `call` | original function call |
| `optimization_info` | information returned by the permutation optimization step |

The exact set of components can differ between LDA-type and QDA-type models.

A compact way to inspect the internal structure is to create a small helper.

```{r}
inspect_model <- function(object) {
  data.frame(
    component = names(object),
    class = vapply(
      object,
      function(x) paste(class(x), collapse = ", "),
      character(1)
    ),
    length = vapply(object, length, integer(1)),
    dim = vapply(
      object,
      function(x) {
        d <- dim(x)
        if (is.null(d)) "" else paste(d, collapse = " x ")
      },
      character(1)
    ),
    row.names = NULL
  )
}
```

```{r}
inspect_model(lda_fit)
```

```{r}
inspect_model(qda_fit)
```

```{r}
inspect_model(joint_qda_fit)
```

The `optimization_info` component stores the result of the `gips` optimization.

```{r}
lda_fit$optimization_info
```

```{r}
qda_fit$optimization_info
```

```{r}
joint_qda_fit$optimization_info
```

When `MAP = TRUE`, this component describes the selected permutation structure.
When `MAP = FALSE`, it contains posterior-probability information used for
weighted averaging.

## LDA diagnostics and visualization

For `gipslda()` objects, standard diagnostic and visualization methods are
available.

```{r}
coef(lda_fit)
```

```{r, fig.width = 6, fig.height = 5}
plot(lda_fit)
```

```{r, fig.width = 6, fig.height = 5}
pairs(lda_fit, type = "std")
```

Depending on the object and method, additional plot types may be available.

```{r, eval = FALSE}
pairs(lda_fit, type = "trellis")
```

## Prediction output

Prediction methods return objects containing predicted classes and posterior
probabilities.

```{r}
names(lda_pred)
names(qda_pred)
names(joint_qda_pred)
```

The most commonly used components are:

| Component | Meaning |
|---|---|
| `class` | predicted class labels |
| `posterior` | posterior class probabilities |
| `x` | discriminant coordinates, when available |

For example:

```{r}
head(lda_pred$class)
head(lda_pred$posterior)
```

## Practical recommendations

Start with `gipslda()` when a common covariance structure across classes is a
reasonable assumption.

```{r}
fit <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)
```

Use `gipsqda()` when each class may have a different covariance structure.

```{r}
fit <- gipsqda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)
```

Use `gipsmultqda()` when class covariance matrices may differ but are expected to
share a common permutation structure.

```{r}
fit <- gipsmultqda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)
```

For small numbers of features, use `optimizer = "BF"`.

```{r}
fit <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "BF"
)
```

For larger numbers of features, use `optimizer = "MH"` and set `max_iter`.

```{r, eval = FALSE}
fit <- gipslda(
  Species ~ .,
  data = train,
  optimizer = "MH",
  max_iter = 1000
)
```

Use `MAP = FALSE` when you want to account for uncertainty in the selected
permutation structure.

```{r}
fit <- gipslda(
  Species ~ .,
  data = train,
  MAP = FALSE,
  optimizer = "BF"
)
```

Use `weighted_avg = TRUE` only for `gipslda()` when you want the alternative
class-size-weighted covariance pooling strategy.

```{r}
fit <- gipslda(
  Species ~ .,
  data = train,
  weighted_avg = TRUE,
  optimizer = "BF"
)
```

## Summary

`gipsDA` provides three main drop-in alternatives to classical discriminant
analysis:

- `gipslda()` regularizes LDA by projecting the pooled covariance matrix onto a
  permutation-invariant structure.
- `gipsqda()` regularizes QDA by projecting each class-specific covariance matrix
  independently.
- `gipsmultqda()` regularizes QDA by estimating one shared permutation structure
  for all class-specific covariance matrices.

The most important arguments are:

| Argument | Main use |
|---|---|
| `MAP` | choose MAP projection or posterior-weighted averaging |
| `optimizer` | choose `"BF"` or `"MH"` |
| `max_iter` | control MH runtime |
| `prior` | set class priors |
| `weighted_avg` | choose the LDA covariance pooling strategy |
| `tol` | numerical tolerance |

For new users, a good starting point is:

```{r}
fit <- gipslda(
  Species ~ .,
  data = train,
  MAP = TRUE,
  optimizer = "BF"
)

pred <- predict(fit, test)

mean(pred$class == test$Species)
```